# Introduction to MoBPS {#sec-introduction}

## Learning Objectives

By the end of this chapter, you will:

- Understand what MoBPS is and what it can do
- Know how to install MoBPS and its dependencies
- Be able to load the package and verify installation
- Understand the basic workflow of a MoBPS simulation

## What is MoBPS?

**MoBPS** (Modular Breeding Program Simulator) is an R package for stochastic simulation of breeding programs. It provides a flexible framework to:

- **Create** realistic founder populations with complex genetic architectures
- **Simulate** multiple generations of breeding actions (selection, mating, culling)
- **Analyze** genetic gain, diversity, inbreeding, and population structure
- **Compare** different breeding strategies
- **Optimize** breeding programs for specific goals

### Why Simulate Breeding Programs?

Breeding program simulation allows you to:

1. **Test strategies** before implementing them in reality
2. **Understand trade-offs** between genetic gain and diversity
3. **Optimize selection** and mating decisions
4. **Predict long-term** outcomes over many generations
5. **Train** future breeders in a risk-free environment

### Key Features

MoBPS offers exceptional flexibility:

- ✅ Works with **both plant and animal breeding**
- ✅ Supports **single or multiple traits**
- ✅ Handles **additive, dominance, and epistatic effects**
- ✅ Simulates **realistic genomic data** or uses real data
- ✅ Integrates with **external prediction software** (BLUP, GBLUP, Bayesian)
- ✅ Provides **rich visualization** and analysis tools
- ✅ Scales from small experiments to **large commercial programs**

## Installation {#sec-installation}

### System Requirements

MoBPS requires:

- **R version 3.0 or higher** (R 4.0+ recommended)
- For large-scale simulations: `RandomFieldsUtils` and `miraculix` (optional but highly recommended)

### Step 1: Platform-Specific Prerequisites

#### Windows Users

Some Windows systems require **Rtools**:

1. Download from: <https://cran.r-project.org/bin/windows/Rtools/>
2. Install the version matching your R version
3. Verify with: `pkgbuild::find_rtools()`

#### Mac Users

Some Mac systems require **gfortran/homebrew**:

1. Download from: <https://mac.r-project.org/tools/>
2. Follow installation instructions for your macOS version

#### Linux Users

Usually no special prerequisites needed. Ensure build tools are installed:

```bash
# Ubuntu/Debian
sudo apt-get install build-essential gfortran

# Fedora/RHEL
sudo dnf install gcc-gfortran
```

### Step 2: Install Performance Packages (Optional but Recommended)

For **significant performance improvements** in large simulations:

```{r}
#| eval: false

# Install devtools if not already installed
if (!require("devtools")) install.packages("devtools")

# Install RandomFieldsUtils (install this FIRST)
devtools::install_github("tpook92/RandomFieldsUtils")

# Then install miraculix (install SECOND)
devtools::install_github("tpook92/miraculix")
```

**Important:**
- Install `RandomFieldsUtils` **before** `miraculix`
- Use compatible versions (both v1.6.0.2 recommended)
- These packages dramatically speed up genomic computations

### Step 3: Install MoBPS

Install the latest MoBPS from GitHub (recommended over CRAN for latest features):

```{r}
#| eval: false

# Install MoBPS from GitHub
devtools::install_github("tpook92/MoBPS", subdir="pkg")
```

**Why GitHub over CRAN?**
- GitHub has the latest features and bug fixes
- Regular updates and improvements
- CRAN version may be outdated (last update: November 2021)

### Step 4: Install Genetic Maps (Optional)

For working with real species data:

```{r}
#| eval: false

# Install MoBPS maps package (includes Ensembl maps)
devtools::install_github("tpook92/MoBPS", subdir="pkg-maps")
```

This package includes genetic maps for common species from Ensembl.

### Step 5: Verify Installation

Load the package and check version:

```{r}
#| eval: false

# Load MoBPS
library(MoBPS)

# Check version
packageVersion("MoBPS")

# Should see something like: '1.13.0' or higher
```

### Additional Packages

MoBPS can integrate with various prediction software. These are **optional** and only needed for specific analyses:

```{r}
#| eval: false

# For genomic prediction
install.packages("rrBLUP")    # GBLUP
install.packages("BGLR")      # Bayesian methods
install.packages("sommer")    # Mixed models

# For visualization
install.packages("ggplot2")   # Advanced plotting
install.packages("viridis")   # Color palettes
```

## Basic Workflow Overview {#sec-workflow}

Every MoBPS simulation follows the same basic structure:

### 1. Create a Founder Population

Use `creating.diploid()` to initialize your population:

```{r}
#| eval: false

population <- creating.diploid(
  nsnp = 1000,          # Number of SNP markers
  nindi = 100,          # Number of individuals
  n.additive = 100,     # Number of QTLs
  mean.target = 100     # Mean breeding value
)
```

### 2. Simulate Breeding Actions

Use `breeding.diploid()` to perform selection, mating, and other actions:

```{r}
#| eval: false

population <- breeding.diploid(
  population,
  selection.size = c(10, 10),  # 10 males, 10 females
  breeding.size = c(50, 50)     # Generate 50 male, 50 female offspring
)
```

### 3. Extract and Analyze Results

Use `get.*()` functions to extract information:

```{r}
#| eval: false

# Get breeding values
bv <- get.bv(population, gen = 2)

# Get phenotypes
pheno <- get.pheno(population, gen = 2)

# Calculate inbreeding
inbreeding <- inbreeding.exp(population, gen = 2)
```

### 4. Visualize Results

Use built-in plotting functions:

```{r}
#| eval: false

# Track genetic gain
bv.development(population)

# PCA plot
get.pca(population, gen = 1:5)

# Kinship development
kinship.development(population, gen = 1:5)
```

## The Population Object

The core of MoBPS is the **population object** (usually called `population` or `pop`):

- It's an **R list** containing all simulation data
- Stores genotypes, phenotypes, pedigrees, trait architectures
- Updated by each `breeding.diploid()` call
- Can be saved/loaded for reproducibility

Think of it as a **complete database** of your breeding program.

## Getting Help

### Within R

```{r}
#| eval: false

# Function documentation
?creating.diploid
?breeding.diploid

# See all MoBPS functions
help(package = "MoBPS")
```

### Online Resources

- **GitHub:** <https://github.com/tpook92/MoBPS>
- **Web Interface:** <https://www.mobps.de>
- **Email Support:** Torsten.pook@wur.nl

### Important Notes

:::{.callout-warning}
## Take Simulation Results with Caution!

MoBPS won't stop you from making biologically impossible settings (e.g., collecting milk records from bulls). Always validate your simulation setup makes biological sense!
:::

:::{.callout-tip}
## Don't Get Overwhelmed!

MoBPS has many parameters, but you only need a few for most simulations. Start simple, then add complexity as needed. Use Ctrl+F to search the documentation for specific features.
:::

## What's Next?

In the next chapter, we'll dive into **core concepts** that are essential for understanding how MoBPS works:

- Individual grouping (generations, databases, cohorts)
- The population list structure
- How traits are represented
- Time flow in simulations

Let's continue to [Chapter 2: Core Concepts](02-core-concepts.qmd)!

## Summary

- MoBPS is a flexible R package for breeding program simulation
- Install from GitHub for latest features
- Optional performance packages (`miraculix`) are highly recommended
- Basic workflow: Create → Breed → Analyze → Visualize
- The population object stores all your simulation data
- Help is available through documentation and direct contact

