# Introduction to MoBPS {#sec-introduction}

## Learning Objectives

By the end of this chapter, you will:

- Understand what MoBPS is and what it can do
- Know how to install MoBPS and its dependencies
- Be able to load the package and verify installation
- Understand the basic workflow of a MoBPS simulation

## What is MoBPS?

**MoBPS** (Modular Breeding Program Simulator) is an R package for stochastic simulation of breeding programs [@pook2020mobps]. It provides a flexible framework to:

- **Create** realistic founder populations with complex genetic architectures
- **Simulate** multiple generations of breeding actions (selection, mating, culling)
- **Analyze** genetic gain, diversity, inbreeding, and population structure
- **Compare** different breeding strategies
- **Optimize** breeding programs for specific goals

### Why Simulate Breeding Programs?

Breeding program simulation allows you to:

1. **Test strategies** before implementing them in reality
2. **Understand trade-offs** between genetic gain and diversity
3. **Optimize selection** and mating decisions
4. **Predict long-term** outcomes over many generations
5. **Train** future breeders in a risk-free environment

### Key Features

MoBPS offers exceptional flexibility:

- ✅ Works with **both plant and animal breeding**
- ✅ Supports **single or multiple traits**
- ✅ Handles **additive, dominance, and epistatic effects**
- ✅ Simulates **realistic genomic data** or uses real data
- ✅ Integrates with **external prediction software** (BLUP, GBLUP, Bayesian)
- ✅ Provides **rich visualization** and analysis tools
- ✅ Scales from small experiments to **large commercial programs**

## Installation {#sec-installation}

### System Requirements

MoBPS requires:

- **R version 3.0 or higher** (R 4.0+ recommended)
- For large-scale simulations: `RandomFieldsUtils` and `miraculix` (optional but highly recommended)

### Step 1: Platform-Specific Prerequisites

#### Windows Users

Some Windows systems require **Rtools**:

1. Download from: <https://cran.r-project.org/bin/windows/Rtools/>
2. Install the version matching your R version
3. Verify with: `pkgbuild::find_rtools()`

#### Mac Users

Mac users need **Xcode Command Line Tools** and **gfortran** (the Mac equivalent of Rtools). The easiest path is via [Homebrew](https://brew.sh).

**Step 1 — Install Xcode Command Line Tools:**

```bash
xcode-select --install
```

A dialog will appear asking you to install. Click "Install" and wait for it to complete. Verify with:

```bash
xcode-select -p
# Should print: /Library/Developer/CommandLineTools
```

**Step 2 — Install Homebrew** (if not already installed):

```bash
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
```

For Apple Silicon (M1/M2/M3) Macs, follow any post-install instructions to add Homebrew to your `PATH`:

```bash
echo 'eval "$(/opt/homebrew/bin/brew shellenv)"' >> ~/.zprofile
eval "$(/opt/homebrew/bin/brew shellenv)"
```

**Step 3 — Install gfortran via Homebrew:**

```bash
brew install gcc
```

This installs `gfortran` (bundled with GCC), which R needs to compile packages with Fortran code (including `miraculix`). Verify:

```bash
gfortran --version
# Should print: GNU Fortran (Homebrew GCC ...) ...
```

**Step 4 — Optional: Install R via Homebrew** (if not already installed):

```bash
brew install --cask r
```

Or download the `.pkg` installer from <https://cran.r-project.org/bin/macosx/> if you prefer.

#### Linux Users

Linux setup varies depending on whether you have admin (sudo) privileges or are working on a shared HPC/cluster system.

---

##### Option 1: Desktop/Server with sudo Access

Install build tools, gfortran, and supporting libraries:

```bash
# Ubuntu / Debian
sudo apt-get update
sudo apt-get install build-essential gfortran liblapack-dev libblas-dev \
  libcurl4-openssl-dev libssl-dev libxml2-dev

# Fedora / RHEL / Rocky Linux / AlmaLinux
sudo dnf install gcc gcc-gfortran gcc-c++ lapack-devel blas-devel \
  libcurl-devel openssl-devel libxml2-devel make

# openSUSE / SLES
sudo zypper install gcc gcc-fortran gcc-c++ lapack-devel blas-devel \
  libcurl-devel libopenssl-devel libxml2-devel
```

Install R (if not already installed):

```bash
# Ubuntu / Debian — add CRAN repo for latest R
sudo apt-get install --no-install-recommends r-base r-base-dev

# Fedora / RHEL
sudo dnf install R
```

---

##### Option 2: HPC / Cluster (module system, no sudo required)

Most HPC systems use **Environment Modules** (`module` command). Load R and the required compiler toolchain before starting R or submitting jobs.

**Check what's available:**

```bash
module avail R
module avail gcc
module avail intel   # Intel compilers also work
```

**Load the modules** (adjust version numbers to what your cluster provides):

```bash
module load R/4.3.1          # or whatever version is available
module load gcc/12.2.0       # provides gfortran
module load openblas         # BLAS/LAPACK for linear algebra performance
```

**Make it persistent** — add to your `~/.bashrc` or `~/.bash_profile` so modules load automatically:

```bash
echo 'module load R/4.3.1' >> ~/.bashrc
echo 'module load gcc/12.2.0' >> ~/.bashrc
source ~/.bashrc
```

**In SLURM job scripts**, load modules at the top of the script:

```bash
#!/bin/bash
#SBATCH --job-name=mobps_sim
#SBATCH --cpus-per-task=4
#SBATCH --mem=16G
#SBATCH --time=02:00:00

module load R/4.3.1
module load gcc/12.2.0

Rscript my_simulation.R
```

---

##### Option 3: HPC / Cluster (no sudo, no module system)

If your cluster lacks the needed modules, use **Conda/Mamba** to install R and compilers entirely in your home directory — no admin privileges needed.

**Install Miniforge** (lightweight Conda with conda-forge defaults):

```bash
# Download and install Miniforge
wget https://github.com/conda-forge/miniforge/releases/latest/download/Miniforge3-Linux-x86_64.sh
bash Miniforge3-Linux-x86_64.sh -b -p ~/miniforge3
~/miniforge3/bin/conda init bash
source ~/.bashrc
```

**Create an environment with R and compilers:**

```bash
conda create -n mobps -c conda-forge r-base r-devtools gfortran_linux-64 \
  openblas lapack -y
conda activate mobps
```

**Install R packages inside the environment** (same as normal — no sudo needed):

```r
install.packages("devtools")
devtools::install_github("tpook92/MoBPS", subdir="pkg")
```

**Set a personal R library path** (useful even outside Conda, e.g. if R is already on the cluster but you want to install your own packages):

```bash
# Add to ~/.bashrc
export R_LIBS_USER=~/R/library
mkdir -p ~/R/library
```

Then in R, verify it's on your search path:

```r
.libPaths()
# ~/R/library should appear first
```

:::{.callout-tip}
## HPC Tip: Ask Your Sysadmin

If MoBPS or its dependencies fail to compile, ask your HPC support team to load or install `gcc` with Fortran support (`gfortran`). Most clusters already have this — it may just need a `module load gcc` to activate it.
:::

### Step 2: Install Performance Packages (Optional but Recommended)

`RandomFieldsUtils` and `miraculix` provide **significant performance improvements** for large simulations by using SIMD CPU instructions and OpenMP multi-threading for genomic matrix operations. For serious simulations — especially on HPC — these are effectively required.

:::{.callout-important}
## Version and Install Order Matter

- Always install **`RandomFieldsUtils` first**, then `miraculix`
- Use **exactly version 1.6.0.2 of both** — this is the version Torsten currently recommends and that is known to work with MoBPS 1.13.x
- Both packages compile C++ code, so a working compiler toolchain is required (see Step 1 above)
:::

Both packages are distributed as source tarballs from the public MoBPS GitHub repository. Download them first, then install from the local files.

#### Get the Tarballs

**Option A — Download via browser:**

Go to <https://github.com/tpook92/MoBPS> and download these two files from the root of the repository:

- `RandomFieldsUtils_1.6.0.2.tar.gz`
- `miraculix_1.6.0.2.tar.gz`

**Option B — Download via curl (Mac / Linux / HPC):**

```bash
mkdir -p ~/mobps-pkgs && cd ~/mobps-pkgs

curl -L -o RandomFieldsUtils_1.6.0.2.tar.gz \
  https://raw.githubusercontent.com/tpook92/MoBPS/master/RandomFieldsUtils_1.6.0.2.tar.gz

curl -L -o miraculix_1.6.0.2.tar.gz \
  https://raw.githubusercontent.com/tpook92/MoBPS/master/miraculix_1.6.0.2.tar.gz
```

**Option C — Download via wget (common on HPC clusters):**

```bash
mkdir -p ~/mobps-pkgs && cd ~/mobps-pkgs

wget https://raw.githubusercontent.com/tpook92/MoBPS/master/RandomFieldsUtils_1.6.0.2.tar.gz
wget https://raw.githubusercontent.com/tpook92/MoBPS/master/miraculix_1.6.0.2.tar.gz
```

---

#### Installing on Windows

Windows users need **Rtools** installed (see Step 1). Then install from the tarballs in R:

```{r}
#| eval: false

# Set path to folder containing the .tar.gz files
pkg_dir <- "C:/Users/yourname/mobps-pkgs"   # adjust to your download folder

install.packages(
  file.path(pkg_dir, "RandomFieldsUtils_1.6.0.2.tar.gz"),
  repos = NULL, type = "source"
)

install.packages(
  file.path(pkg_dir, "miraculix_1.6.0.2.tar.gz"),
  repos = NULL, type = "source"
)
```

Or equivalently from a terminal (Command Prompt or PowerShell):

```bash
R CMD INSTALL RandomFieldsUtils_1.6.0.2.tar.gz
R CMD INSTALL miraculix_1.6.0.2.tar.gz
```

---

#### Installing on Mac

Mac's default Apple Clang compiler **does not support OpenMP**, which both packages require. You must first configure R to use GCC (installed via Homebrew in Step 1) and install `libomp`.

**Step 1 — Install OpenMP support via Homebrew:**

```bash
brew install libomp
```

:::{.callout-note}
## About the Homebrew libomp warning

Homebrew will warn that `libomp` was not symlinked and may suggest setting `LDFLAGS` and `CPPFLAGS` in your shell profile. You can **ignore this warning** for our purposes.

`libomp` is Apple Clang's OpenMP library. The warning exists because Homebrew intentionally doesn't symlink it to avoid conflicts with Xcode. However, the `~/.R/Makevars` setup below routes R through **Homebrew GCC**, which uses its own OpenMP library (`libgomp`) — it never looks for `libomp` at all. The `LDFLAGS`/`CPPFLAGS` Homebrew suggests are only needed if you were using Apple Clang with OpenMP, which we are not.
:::

**Step 2 — Configure R to use Homebrew GCC:**

Create (or edit) the file `~/.R/Makevars` to override the default compiler:

```bash
mkdir -p ~/.R
```

Then create `~/.R/Makevars` with the following content (adjust GCC version number to match what `brew install gcc` installed — check with `ls /opt/homebrew/bin/gcc*`):

```makefile
# ~/.R/Makevars
CC=/opt/homebrew/bin/gcc-14
CXX=/opt/homebrew/bin/g++-14
CXX11=/opt/homebrew/bin/g++-14
CXX14=/opt/homebrew/bin/g++-14
CXX17=/opt/homebrew/bin/g++-14
FC=/opt/homebrew/bin/gfortran-14
F77=/opt/homebrew/bin/gfortran-14
OBJC=/opt/homebrew/bin/gcc-14

# OpenMP flags
SHLIB_OPENMP_CFLAGS=-fopenmp
SHLIB_OPENMP_CXXFLAGS=-fopenmp
SHLIB_OPENMP_FFLAGS=-fopenmp
```

:::{.callout-tip}
## Finding your GCC version number

```bash
ls /opt/homebrew/bin/gcc-*
# e.g. /opt/homebrew/bin/gcc-14  → use 14
```
:::

**Step 3 — Install from tarballs in R:**

```{r}
#| eval: false

pkg_dir <- "~/mobps-pkgs"   # adjust to your path

install.packages(
  file.path(pkg_dir, "RandomFieldsUtils_1.6.0.2.tar.gz"),
  repos = NULL, type = "source"
)

install.packages(
  file.path(pkg_dir, "miraculix_1.6.0.2.tar.gz"),
  repos = NULL, type = "source"
)
```

Or from the terminal:

```bash
cd ~/mobps-pkgs
R CMD INSTALL RandomFieldsUtils_1.6.0.2.tar.gz
R CMD INSTALL miraculix_1.6.0.2.tar.gz
```

:::{.callout-note}
## Apple Silicon (M1/M2/M3) Note

The packages use x86 SIMD instructions (AVX/SSE) that are not natively available on Apple Silicon. The `configure.ac` in both packages detects the architecture and falls back to a non-SIMD build automatically — you will still get a working install, just without the maximum SIMD speedup. The OpenMP parallelism still works and provides meaningful gains.
:::

---

#### Installing on Linux with sudo

```bash
# Ubuntu/Debian — ensure build tools are present (see Step 1)
sudo apt-get install libgomp1

# Fedora/RHEL
sudo dnf install libgomp
```

Then install from tarballs in R:

```{r}
#| eval: false

pkg_dir <- "~/mobps-pkgs"   # adjust to your path

install.packages(
  file.path(pkg_dir, "RandomFieldsUtils_1.6.0.2.tar.gz"),
  repos = NULL, type = "source"
)

install.packages(
  file.path(pkg_dir, "miraculix_1.6.0.2.tar.gz"),
  repos = NULL, type = "source"
)
```

Or from bash directly:

```bash
cd ~/mobps-pkgs
R CMD INSTALL RandomFieldsUtils_1.6.0.2.tar.gz
R CMD INSTALL miraculix_1.6.0.2.tar.gz
```

---

#### Installing on HPC / Cluster (module system)

Load the required modules first, then install into your personal R library:

```bash
module load R/4.3.1
module load gcc/12.2.0      # must include gfortran

# Ensure personal library exists
mkdir -p ~/R/library
export R_LIBS_USER=~/R/library
```

Then install from the tarballs:

```bash
cd ~/mobps-pkgs

R CMD INSTALL --library=~/R/library RandomFieldsUtils_1.6.0.2.tar.gz
R CMD INSTALL --library=~/R/library miraculix_1.6.0.2.tar.gz
```

Add the library path to your `~/.bashrc` so R always finds these packages:

```bash
echo 'export R_LIBS_USER=~/R/library' >> ~/.bashrc
```

And add the same to the top of your SLURM job scripts:

```bash
#!/bin/bash
#SBATCH --job-name=mobps_sim
#SBATCH --cpus-per-task=8
#SBATCH --mem=32G

module load R/4.3.1
module load gcc/12.2.0

export R_LIBS_USER=~/R/library
Rscript my_simulation.R
```

---

#### Installing on HPC without sudo or modules (Conda)

If your cluster uses Conda/Mamba (see Option 3 in Step 1), the compilers and OpenMP support are already bundled in the environment. Just activate it and install:

```bash
conda activate mobps

cd ~/mobps-pkgs
R CMD INSTALL RandomFieldsUtils_1.6.0.2.tar.gz
R CMD INSTALL miraculix_1.6.0.2.tar.gz
```

---

#### Verify the Installation

In R, confirm both packages load without errors:

```{r}
#| eval: false

library(RandomFieldsUtils)
library(miraculix)

# Check versions
packageVersion("RandomFieldsUtils")  # should be 1.6.0.2
packageVersion("miraculix")          # should be 1.6.0.2
```

If `miraculix` loads successfully, MoBPS will automatically detect and use it for faster genomic computations — no extra configuration needed.

### Step 3: Install MoBPS

Install the latest MoBPS from GitHub (recommended over CRAN for latest features):

```{r}
#| eval: false

# Install MoBPS from GitHub
devtools::install_github("tpook92/MoBPS", subdir="pkg")
```

**Why GitHub over CRAN?**
- GitHub has the latest features and bug fixes
- Regular updates and improvements
- CRAN version may be outdated (last update: November 2021)

### Step 4: Install Genetic Maps (Optional)

For working with real species data:

```{r}
#| eval: false

# Install MoBPS maps package (includes Ensembl maps)
devtools::install_github("tpook92/MoBPS", subdir="pkg-maps")
```

This package includes genetic maps for common species from Ensembl.

### Step 5: Verify Installation

Load the package and check version:

```{r}
#| eval: false

# Load MoBPS
library(MoBPS)

# Check version
packageVersion("MoBPS")

# Should see something like: '1.13.0' or higher
```

### Additional Packages

MoBPS can integrate with various prediction software. These are **optional** and only needed for specific analyses:

```{r}
#| eval: false

# For genomic prediction
install.packages("rrBLUP")    # GBLUP
install.packages("BGLR")      # Bayesian methods
install.packages("sommer")    # Mixed models

# For visualization
install.packages("ggplot2")   # Advanced plotting
install.packages("viridis")   # Color palettes
```

## Basic Workflow Overview {#sec-workflow}

Every MoBPS simulation follows the same basic structure:

### 1. Create a Founder Population

Use `creating.diploid()` to initialize your population:

```{r}
#| eval: false

population <- creating.diploid(
  nsnp = 1000,          # Number of SNP markers
  nindi = 100,          # Number of individuals
  n.additive = 100,     # Number of QTLs
  mean.target = 100     # Mean breeding value
)
```

### 2. Simulate Breeding Actions

Use `breeding.diploid()` to perform selection, mating, and other actions:

```{r}
#| eval: false

population <- breeding.diploid(
  population,
  selection.size = c(10, 10),  # 10 males, 10 females
  breeding.size = c(50, 50)     # Generate 50 male, 50 female offspring
)
```

```{r}
#| eval: false

# phenotype generation 2
population <- breeding.diploid(
  population,
  phenotyping.gen = 2
)
```

### 3. Extract and Analyze Results

Use `get.*()` functions to extract information:

```{r}
#| eval: false

# Get breeding values of gen 2
bv <- get.bv(population, gen = 2)

# Get phenotypes of gen 2
pheno <- get.pheno(population, gen = 2)

# Calculate inbreeding for gen 2
inbreeding <- inbreeding.exp(population, gen = 2)
```

### 4. Visualize Results

Use built-in plotting functions:

```{r}
#| eval: false

# Track genetic gain
bv.development(population)

# PCA plot
get.pca(population, gen = 1:2)

# Kinship development
kinship.development(population, gen = 1:2)
```

## The Population Object

The core of MoBPS is the **population object** (usually called `population` or `pop`):

- It's an **R list** containing all simulation data
- Stores genotypes, phenotypes, pedigrees, trait architectures
- Updated by each `breeding.diploid()` call
- Can be saved/loaded for reproducibility

Think of it as a **complete database** of your breeding program.

## Getting Help

### Within R

```{r}
#| eval: false

# Function documentation
?creating.diploid
?breeding.diploid

# See all MoBPS functions
help(package = "MoBPS")
```

### Online Resources

- **GitHub:** <https://github.com/tpook92/MoBPS>
- **Web Interface:** <https://www.mobps.de>
- **Email Support:** Torsten.pook@wur.nl

### Important Notes

:::{.callout-warning}
## Take Simulation Results with Caution!

MoBPS won't stop you from making biologically impossible settings (e.g., collecting milk records from bulls). Always validate your simulation setup makes biological sense!
:::

:::{.callout-tip}
## Don't Get Overwhelmed!

MoBPS has many parameters, but you only need a few for most simulations. Start simple, then add complexity as needed. Use Ctrl+F to search the documentation for specific features.
:::

## What's Next?

In the next chapter, we'll dive into **core concepts** that are essential for understanding how MoBPS works:

- Individual grouping (generations, databases, cohorts)
- The population list structure
- How traits are represented
- Time flow in simulations

Let's continue to [Chapter 2: Core Concepts](02-core-concepts.qmd)!

## Summary

- MoBPS is a flexible R package for breeding program simulation
- Install from GitHub for latest features
- Optional performance packages (`miraculix`) are highly recommended
- Basic workflow: Create → Breed → Analyze → Visualize
- The population object stores all your simulation data
- Help is available through documentation and direct contact

