# Core Concepts {#sec-core-concepts}

## Learning Objectives

By the end of this chapter, you will understand:

- The gen/database/cohorts system for grouping individuals
- How sex is handled in MoBPS
- The structure of the population object
- How time flows through generations
- Key terminology and concepts

## Individual Grouping {#sec-grouping}

One of the biggest challenges in breeding simulation is having the flexibility to perform operations on **specific groups** of individuals. MoBPS provides three powerful ways to select groups:

1. **Generations** (`gen`) - Select all individuals from specific generation(s)
2. **Cohorts** (`cohorts`) - Select named groups with specific characteristics
3. **Database** (`database`) - Precise selection by generation, sex, and individual range

### Understanding Generations (`gen`)

Every time you create offspring with `breeding.diploid()`, they are assigned to a **new generation**. Generations are numbered sequentially starting from 1.

```{r}
#| eval: false

# Create founder population (generation 1)
population <- creating.diploid(
                nsnp = 1000,      # 1000 total SNP
				nindi = 100,      # 100 individuals in First Cohort, 50 Males + 50 Females
                mean.target = 0,  # target mean TBV
				var.target = 1,   # target variance of TBV
				n.additive = 100, # 100 QTL
				name.cohort = "Founders"
				)

# Create generation 2
# select 10 males and 10 females from gen 1
# mate them to create 50 new males and 50 new females
population <- breeding.diploid(population,
                               selection.size = c(10, 10), # select 50 males and 50 females from past cohorts
                               breeding.size = c(50, 50)   # generate new cohort, 50 males and 50 females
							   )

# extract True Breeding Value (BV) from population list - returns a basic vector
bv_gen2 <- get.bv(population, gen = 2)

# extract True Breeding Value (BV) from both generations
bv_multi <- get.bv(population, gen = 1:2)  # Generations 1 and 2
```

**Key points:**

- Generations track **when** individuals were born
- Useful for age-structured populations
- Easy to analyze trends over time

### Named Groups: Cohorts {#sec-cohorts}

**Cohorts** are named groups of individuals you define. They're incredibly useful for tracking:

- Different selection lines (e.g., "HighYield", "LowFat")
- Breeding groups (e.g., "NucleusHerd", "CommercialLine")
- Treatment groups (e.g., "Tested", "Controls")

```{r}
#| eval: false

# Create offspring and assign to named cohort
population <- breeding.diploid(
  population,
  selection.size = c(10, 10),  # select 10 males and 10 females
  breeding.size = c(50, 50),   # generate 50 males and 50 females
  name.cohort = "SelectedLine"  # Give this group (cohort) a name
)

# Later, extract results such as the TBV by name
bv_selected <- get.bv(population, cohorts = "SelectedLine")

# Select multiple cohorts
multi_cohorts <- get.bv(population,
                         cohorts = c("Cohort_1", "SelectedLine"))
```

**Best practices:**

- Use descriptive names: "TopSires", "TestGroup1", "Line_A"
- Keep naming consistent across simulations
- Use cohorts when generation number alone isn't enough

### Precise Selection: Database {#sec-database}

The **database** parameter gives you surgical precision. It's a matrix where each row specifies:

1. **Generation** number
2. **Sex** (1 = male, 2 = female)
3. **First individual** to include (optional)
4. **Last individual** to include (optional)

```{r}
#| eval: false

# create 1 by 4 matrix with info from above
database <- matrix(c(1, 1, 1, 20), ncol = 4)

# Extract TBV of males 1-20 from generation 1
males <- get.bv(population, database = database)

# Select a specific range (4-column: gen, sex, first, last)
db_sub <- rbind(
  c(1, 1, 1, 20),    # Males 1-20 from gen 1
  c(1, 2, 5, 15)     # Females 5-15 from gen 1
)
bv_db_sub <- get.bv(population, database = db_sub)
```

You may want to select all animals from a gen + sex combination, we can do this with a 1x2 matrix using `database`.

```{r}
#| eval: false

# Select ALL individuals of a sex/gen: pass only 2 columns (gen, sex)
# MoBPS auto-fills start=1 and end=N for that group
db_all_males <- matrix(c(2, 1), ncol = 2)  # All males from gen 2
bv_all_males <- get.bv(population, database = db_all_males)
```

If we want to select 20 males from gen 1 and all males from gen 2,
we can use the following code. 

```{r}
#| eval: false

# Combining full-range rows with specific-range rows requires explicit counts
# because rbind() with mixed column counts will pad with NA (which breaks things)
db_combined <- rbind(
  c(1, 1, 1, 20),                                    # Males 1-20 from gen 1
  c(2, 1, 1, population$info$size[2, 1])             # All males from gen 2
)
bv_combined <- get.bv(population, database = db_combined)
```

> Remember: Each line goes from 1 to n (counting restarts each gen + sex)

Run the following code to make this clear. We can nest `get.` functions
like this to extract data from all cohorts if needed. 

```{r}
#| eval: false

get.database(population, cohorts=get.cohorts(population))
```

**When to use database:**

- Need specific individual ranges
- Complex selection criteria
- Combining multiple precise selections

### Combining Extraction Methods

You can mix and match with `gen`, `database`, and `cohorts` if needed for some reason as follows:

```{r}
#| eval: false

# add 3 generations for example
population <- breeding.diploid(population, selection.size = c(10, 10), breeding.size = c(50, 50))
population <- breeding.diploid(population, selection.size = c(10, 10), breeding.size = c(50, 50))
population <- breeding.diploid(population, selection.size = c(10, 10), breeding.size = c(50, 50))

# Select from generations 4-5, specific males from gen 3, AND a cohort
database <- matrix(c(3, 1, 21, 50), ncol = 4)

# extract TBV vector based on gen + database + cohorts
bv <- get.bv(population,
             gen = 4:5,              # All of gen 4 & 5
             database = database,     # Males 21-50 from gen 3
             cohorts = "Founders")    # Plus the Founders cohort
```

This gives you incredible flexibility to work with exactly the individuals you need!

## Sex Handling {#sec-sex}

MoBPS has a flexible approach to sex:

### Traditional Two-Sex Systems

By default, individuals are assigned **male = 1** or **female = 2**:

Using `sex.quota` argument to set the **percent of females**:

```{r}
#| eval: false

# Control sex ratio in founders
population <- creating.diploid(
  nsnp = 1000,
  nindi = 100,
  sex.quota = 0.2  # 20% female (or 20 females in this example)
)
```

Using `sex.s` argument:

```{r}
#| eval: false

# Or specify exactly
population <- creating.diploid(
  nsnp = 1000,
  nindi = 100,
  sex.s = c(rep(1, 40), rep(2, 60))  # 40 males, 60 females
)
```

### Flexible Sex Usage

**Important:** Sex assignments are **not binding** for breeding operations!

- An individual stored as "female" can be used as father
- Useful for plant breeding where sex may not be fixed
- Useful for modeling hermaphrodites or aquaculture

### One-Sex Mode

For organisms without meaningful sex distinctions:

```{r}
#| eval: false

# Deactivate two-sex system
population <- creating.diploid(
  nsnp = 1000,
  nindi = 100,
  one.sex.mode = TRUE  # All individuals in "sex 1"
)
```

This automatically adjusts `breeding.size`, `selection.size`, etc. to work with a single group.

### Using Sex as Structure

Even in plants, you can use "sex" to organize populations:

- Sex 1 = Gene pool A
- Sex 2 = Gene pool B

This provides convenient structure for tracking different groups!

## The Population Object {#sec-population-object}

The population object is the **heart of MoBPS**. It's an R list containing everything about your breeding program.

### Main Components

The population object has two major sections:

```{r}
#| eval: false

# Examine top-level structure
names(population)
# [1] "info"     "breeding"

# General information
names(population$info)
# QTL effects, genetic maps, trait info, etc.

# $breeding is a nested numeric list: [[gen]][[sex]][[individual]]
# names() returns NULL because it uses integer indices, not names.
# To inspect it, use:

length(population$breeding)         # number of generations stored
population$info$size                # matrix of individual counts: rows=gen, cols=sex (1=M, 2=F)
str(population$breeding, max.level = 2)  # overview of nesting depth
```

### What's Stored?

**In `$info` (population-level):**

- Genetic map (chromosome structure, marker positions)
- QTL effects and locations
- Trait names and architectures
- Correlation structures
- Fixed effects

**In `$breeding` (individual-level):**

- Genotypes/haplotypes for each individual
- Breeding values (true genetic values)
- Phenotypes (observed values)
- Pedigree information
- Genotyping status
- Cohort assignments

### Accessing the Population Object

While you *can* access elements directly, it's better to use `get.*()` functions:

```{r}
#| eval: false

# DON'T DO THIS (fragile, complex):
# would access data on animal 6 from generation 2 males
bv <- population$breeding[[2]][[1]][[6]]

# DO THIS INSTEAD (clear, robust):
bv <- get.bv(population, gen = 2)
```

**Available getter functions:**

- `get.bv()` - Breeding values
- `get.pheno()` - Phenotypes
- `get.geno()` - Genotypes
- `get.pedigree()` - Pedigree
- `get.map()` - Genetic map
- Many more! (See [Chapter 17](#sec-function-reference))

## Classes: Advanced Grouping {#sec-classes}

**Classes** provide another layer of organization:

```{r}
#| eval: false

# Assign individuals to class 1
population <- set.class(population,
                        gen = 2,
                        new.class = 1)

# Only phenotype class 1
population <- breeding.diploid(
  population,
  phenotyping.gen = 2,
  phenotyping.class = 1  # Only these get phenotyped
)
```

**Special classes:**

- Class 0 (default): Normal active individuals
- Class -1: Culled/dead animals (automatically assigned)
- Class 1+: User-defined groups

Use classes for:

- Active vs. culled animals
- Test groups vs. controls
- Multiple herds/flocks with different management

## Time and Generation Flow {#sec-time-flow}

Understanding how time works in MoBPS is crucial:

### Sequential Generations

Each `breeding.diploid()` call creates the **next generation**:

```{r}
#| eval: false

# Gen 1: Founders
pop <- creating.diploid(nsnp = 1000, nindi = 100)

# Gen 2: First offspring
pop <- breeding.diploid(pop,
                        selection.size = c(10, 10),
                        breeding.size = c(50, 50))

# Gen 3: Second offspring generation
pop <- breeding.diploid(pop,
                        selection.size = c(10, 10),
                        breeding.size = c(50, 50))
```

Generation numbers are **automatic** and **sequential**.

### Overlapping Generations

You can have overlapping generations by selecting parents from multiple generations:

```{r}
#| eval: false

# Use individuals from generation 2 AND 3 as parents
pop <- breeding.diploid(pop,
                        selection.m.gen = 2,      # select Males from gen 2
                        selection.f.gen = 3,      # select Females from gen 3
                        breeding.size = c(50, 50)) # generate 50 new males and 50 new females
```

This creates **generation 4** from a mix of gen 2 and 3 parents, mirroring the overlapping-generation response framework described by Hill [@hill1974prediction].

### Age Structure

MoBPS doesn't explicitly track age, but you can model it:

- Use **generations** as age cohorts
- Use **cohorts** to track birth years
- Select parents from specific generations to control age

See Section 6.20 in the full manual for detailed age structure examples.

## Genetic Pools and Crossbreeding {#sec-pools}

**Founder pools** track the origin of genome segments:

```{r}
#| eval: false

# Create two founder populations
pop1 <- creating.diploid(nsnp = 1000, nindi = 50,
                         founder.pool = 1)  # Pool 1

pop2 <- creating.diploid(nsnp = 1000, nindi = 50,
                         founder.pool = 2)  # Pool 2
```

**Why use pools?**

- Track breed composition in crossbreeding
- Assign breed-specific QTL effects
- Analyze admixture and introgression
- Model heterosis and breed complementarity

You can later query which parts of the genome came from which pool using `get.pool()`.

## Key Terminology Recap

| Term | Definition |
|------|------------|
| **Generation** | Time point when individuals were born (sequential) |
| **Cohort** | Named group of individuals with similar characteristics |
| **Database** | Precise selection by gen, sex, and individual range |
| **Class** | Category for management actions (0=active, -1=culled, 1+=custom) |
| **Sex** | Male (1) or female (2), flexibly used |
| **Pool** | Founder population origin for tracking breed composition |
| **Population object** | R list containing all simulation data |

## Practical Tips

:::{.callout-tip}
## Start Simple
Don't try to use all features at once! Start with just `gen` for selecting individuals, then add cohorts and classes as needed.
:::

:::{.callout-tip}
## Naming Conventions
Use consistent naming:
- Cohorts: "Line_A", "Test_2024", "HighYield"
- Variables: `pop` or `population` for the population object
- Clear generation references in comments
:::

:::{.callout-warning}
## Don't Lose Your Population Object!
Always save important population objects:
```{r}
#| eval: false
saveRDS(population, "my_population_gen10.rds")
# Later:
population <- readRDS("my_population_gen10.rds")
```
:::

## Summary

- **Three selection systems:** gen (generations), cohorts (named groups), database (precise)
- **Flexible sex:** Can be biological sex, gene pools, or organizational structure
- **Population object:** Central data structure storing all simulation information
- **Classes:** Additional grouping for management (0=active, -1=culled, custom)
- **Generations flow sequentially:** Each `breeding.diploid()` creates the next generation
- **Pools track origins:** Useful for crossbreeding and admixture

## What's Next?

Now that you understand the core concepts, let's put them into practice!

In [Chapter 3: Your First Simulation](03-first-simulation.qmd), you'll create a complete breeding program from start to finish.

