# Your First Simulation {#sec-first-simulation}

## Learning Objectives

By the end of this chapter, you will be able to:

- Create a complete breeding program simulation from scratch
- Phenotype and genotype individuals
- Perform genomic prediction
- Select superior individuals
- Generate offspring from selected parents
- Analyze genetic gain and inbreeding

## The Complete Workflow

Let's simulate a realistic breeding program with:

- 100 founder animals
- Genomic selection based on GBLUP
- 10 generations of selection
- Analysis of genetic gain and diversity

## Step 1: Load MoBPS

```{r}
#| eval: false

# Load the package
library(MoBPS)

# Optional: Set seed for reproducibility
set.seed(12345)
```

## Step 2: Create Founder Population {#sec-create-founders}

```{r}
#| eval: false

# Create a founder population
population <- creating.diploid(
  nsnp = 10000,                # 10,000 SNP markers
  nindi = 100,                 # 100 individuals (50 male, 50 female)
  chr.nr = 5,                  # 5 chromosomes
  chromosome.length = 2,       # 2 Morgan each (10M total)
  n.additive = 100,            # 100 additive QTLs
  n.dominant = 20,             # 20 dominant QTLs
  share.genotyped = 1,         # 100% genotyped
  var.target = 1,              # Genetic variance = 1
  name.cohort = "Founders"     # Name this cohort
)
```

**What did we just create?**

- A population with realistic genomic structure (5 chromosomes)
- 100 animals with 10,000 SNPs each
- A complex trait with both additive and dominance effects
- All individuals are genotyped (ready for genomic selection)

## Step 3: Generate Phenotypes {#sec-phenotype}

```{r}
#| eval: false

# Phenotype all individuals in generation 1
population <- breeding.diploid(
  population,
  phenotyping = "all",         # Phenotype everyone
  phenotyping.gen = 1,         # From generation 1
  heritability = 0.5           # h² = 0.5 (moderate heritability)
)
```

**What happened?**

- Phenotypes were simulated for all 100 founders
- Residual variance was automatically calculated to achieve h² = 0.5
- Phenotypes = True Breeding Value + Environmental Effect

## Step 4: Genomic Prediction {#sec-prediction}

```{r}
#| eval: false

# Perform genomic prediction (GBLUP)
population <- breeding.diploid(
  population,
  bve = TRUE,                  # Calculate breeding value estimates
  bve.gen = 1                  # Use generation 1 as training set
)
```

**What happened?**

- A genomic relationship matrix (G) was calculated
- Breeding values were estimated using the internal GBLUP
- Each individual now has an estimated breeding value (EBV)

:::{.callout-note}
## Internal BVE Method

The `bve = TRUE` uses MoBPS's internal GBLUP assuming **known heritability**. For more realistic scenarios, use external software like `rrBLUP`, `BGLR`, or `BLUPF90` (covered in [Chapter 13](#sec-genomic-prediction)).
:::

## Step 5: Selection of Parents {#sec-selection}

```{r}
#| eval: false

# Select top individuals as parents
population <- breeding.diploid(
  population,
  selection.size = c(10, 10),         # 10 males, 10 females
  selection.criteria = "bve",         # Select on estimated BV
  selection.m.cohorts = "Founders",   # Males from Founders
  selection.f.cohorts = "Founders",   # Females from Founders
  breeding.size = c(50, 50),          # Generate 50M + 50F offspring
  name.cohort = "Generation_2"        # Name the offspring
)
```

**What happened?**

- The top 10 males (by EBV) were selected as sires
- The top 10 females (by EBV) were selected as dams
- 100 offspring (50 male, 50 female) were created via random mating
- Selection intensity: 10% of males, 10% of females

## Step 6: Repeat for Multiple Generations {#sec-repeat}

```{r}
#| eval: false

# Simulate 10 generations
for (gen in 2:10) {

  # Phenotype the current generation
  population <- breeding.diploid(
    population,
    phenotyping = "all",
    phenotyping.gen = gen,
    heritability = 0.5
  )

  # Update genomic predictions using all data
  population <- breeding.diploid(
    population,
    bve = TRUE,
    bve.gen = 1:gen                   # Cumulative training set
  )

  # Select and breed
  population <- breeding.diploid(
    population,
    selection.size = c(10, 10),
    selection.criteria = "bve",
    selection.gen = gen,               # Select from current generation
    breeding.size = c(50, 50),
    name.cohort = paste0("Generation_", gen+1)
  )
}
```

**What's happening in the loop?**

1. Each generation is phenotyped
2. Predictions are updated with cumulative data (generations 1 to current)
3. Top 10 males and 10 females are selected
4. 100 offspring are produced

## Step 7: Analyze Results {#sec-analyze}

### Genetic Gain

```{r}
#| eval: false

# Plot breeding value trends
bv.development(population)
```

This creates a plot showing:
- Mean breeding value by generation
- Genetic gain over time
- Standard deviation within generation

### Extract Breeding Values

```{r}
#| eval: false

# Get true breeding values for all generations
bv_all <- get.bv(population, gen = 1:11)

# Calculate mean BV per generation
mean_bv <- apply(bv_all, 2, mean)
mean_bv

# Calculate genetic gain from generation 1 to 11
genetic_gain <- mean_bv[11] - mean_bv[1]
cat("Total genetic gain:", round(genetic_gain, 3), "genetic SD\n")
```

### Inbreeding Analysis

```{r}
#| eval: false

# Calculate genomic inbreeding for generation 11
inbreeding_gen11 <- inbreeding.emp(population, gen = 11)
mean(inbreeding_gen11)

# Track inbreeding across generations
inbreeding_trend <- numeric(11)
for (gen in 1:11) {
  inbreeding_trend[gen] <- mean(inbreeding.emp(population, gen = gen))
}

# Plot inbreeding trend
plot(1:11, inbreeding_trend,
     type = "b",
     xlab = "Generation",
     ylab = "Mean Inbreeding Coefficient",
     main = "Inbreeding Over Time",
     col = "red", lwd = 2)
```

### Kinship Between Individuals

```{r}
#| eval: false

# Calculate kinship matrix for generation 11
kinship_matrix <- kinship.emp(population, gen = 11)

# Average kinship (off-diagonal elements)
n <- nrow(kinship_matrix)
avg_kinship <- sum(kinship_matrix - diag(diag(kinship_matrix))) / (n * (n - 1))
cat("Average kinship in gen 11:", round(avg_kinship, 4), "\n")
```

### Population Structure (PCA)

```{r}
#| eval: false

# PCA of generations 1, 5, and 11
get.pca(population, gen = c(1, 5, 11))
```

This creates a PCA plot showing how the population has diverged from founders.

## Complete Script

Here's the entire simulation in one script:

```{r}
#| eval: false

library(MoBPS)
set.seed(12345)

# 1. Create founders
population <- creating.diploid(
  nsnp = 10000, nindi = 100, chr.nr = 5, chromosome.length = 2,
  n.additive = 100, n.dominant = 20, share.genotyped = 1,
  var.target = 1, name.cohort = "Founders"
)

# 2. Phenotype founders
population <- breeding.diploid(
  population, phenotyping = "all", phenotyping.gen = 1, heritability = 0.5
)

# 3. Initial genomic prediction
population <- breeding.diploid(population, bve = TRUE, bve.gen = 1)

# 4. Create first offspring generation
population <- breeding.diploid(
  population, selection.size = c(10, 10), selection.criteria = "bve",
  selection.m.cohorts = "Founders", selection.f.cohorts = "Founders",
  breeding.size = c(50, 50), name.cohort = "Generation_2"
)

# 5. Repeat for 9 more generations
for (gen in 2:10) {
  population <- breeding.diploid(
    population, phenotyping = "all", phenotyping.gen = gen, heritability = 0.5
  )
  population <- breeding.diploid(population, bve = TRUE, bve.gen = 1:gen)
  population <- breeding.diploid(
    population, selection.size = c(10, 10), selection.criteria = "bve",
    selection.gen = gen, breeding.size = c(50, 50),
    name.cohort = paste0("Generation_", gen+1)
  )
}

# 6. Analyze results
bv.development(population)
bv_all <- get.bv(population, gen = 1:11)
mean_bv <- apply(bv_all, 2, mean)
cat("Genetic gain:", round(mean_bv[11] - mean_bv[1], 3), "\n")

inbreeding_gen11 <- inbreeding.emp(population, gen = 11)
cat("Mean inbreeding gen 11:", round(mean(inbreeding_gen11), 4), "\n")
```

## Understanding the Results

### Expected Outcomes

After 10 generations of selection with these parameters, you should see:

- **Genetic gain:** ~2-4 genetic standard deviations
- **Inbreeding:** ~5-15% (due to small effective population size)
- **Selection response:** Decreasing per generation (plateauing)
- **Prediction accuracy:** Improving with cumulative training data

### Why These Results?

1. **Strong selection** (10% selected) = high genetic gain
2. **Small Ne** (10 males) = rapid inbreeding accumulation
3. **Moderate h²** (0.5) = good response to selection
4. **Finite genome** = some exhaustion of genetic variance

## Variations to Try

Modify the simulation to explore different scenarios:

### Higher Selection Intensity

```{r}
#| eval: false

# More extreme selection
selection.size = c(5, 5)  # Top 5% of each sex
```

### Larger Population

```{r}
#| eval: false

# Reduce inbreeding
nindi = 500,                # 500 founders
selection.size = c(25, 25),  # But still select top 10%
breeding.size = c(250, 250)  # Maintain pop size
```

### Multiple Traits

```{r}
#| eval: false

# Add a second trait
n.additive = c(100, 80),           # Trait 1 & 2
trait.cor = matrix(c(1, -0.3,      # Negative correlation
                      -0.3, 1), nrow = 2)
```

### Different Mating Strategies

```{r}
#| eval: false

# Avoid inbreeding
avoid.mating.halfsib = TRUE,
avoid.mating.fullsib = TRUE
```

## Common Issues and Solutions

:::{.callout-warning}
## "Not enough males/females"

**Problem:** Selection.size too large for available individuals.

**Solution:** Check population size:
```{r}
#| eval: false
get.size(population, gen = 2)
```
Ensure `selection.size` doesn't exceed population size.
:::

:::{.callout-warning}
## "No phenotypes available"

**Problem:** Forgot to phenotype before BVE.

**Solution:** Always phenotype before prediction:
```{r}
#| eval: false
population <- breeding.diploid(population,
                                phenotyping = "all",
                                heritability = 0.5)
```
:::

:::{.callout-tip}
## Speeding Up Simulations

For large-scale simulations:
- Reduce `nsnp` (1000 SNPs often sufficient)
- Use `miraculix` package for fast matrix operations
- Set `copy.breeding.values = FALSE` to reduce memory
:::

## Summary

In this chapter, you:

- ✅ Created a complete breeding program simulation
- ✅ Implemented genomic selection with GBLUP
- ✅ Tracked genetic gain over 10 generations
- ✅ Analyzed inbreeding and population structure
- ✅ Learned the core MoBPS workflow

## What's Next?

Now that you can run a basic simulation, let's dive deeper into each component:

- **Part 2:** Learn how to create more realistic founder populations with LD structure, import real data, and design complex trait architectures
- **[Chapter 4](#sec-creating-populations):** Advanced population creation
- **[Chapter 5](#sec-trait-architecture):** Sophisticated trait modeling

Continue to [Chapter 4: Creating Populations](04-creating-populations.qmd)!

