# Creating Founder Populations {#sec-creating-populations}

## Learning Objectives

By the end of this chapter, you will be able to:

- Generate simulated genomic data with different patterns
- Import real genomic data (VCF, PLINK formats)
- Set up genetic maps and chromosome structures
- Control allele frequencies and genetic diversity
- Create multiple breeds/populations

## Overview of `creating.diploid()`

The `creating.diploid()` function is your starting point for every MoBPS simulation. It creates the **founder population** with:

- Genomic data (markers/haplotypes)
- Genetic map (chromosomes, positions)
- Trait architecture (QTLs, effects)
- Population structure (cohorts, sex, pools)

## Generating Simulated Data {#sec-simulated-data}

### Basic Random Population

The simplest approach: generate random genotypes.

```{r}
#| eval: false

# Create a simple founder population
population <- creating.diploid(
  nsnp       = 5000,    # 5,000 SNP markers
  nindi      = 200,     # 200 individuals (100 each sex)
  n.additive = 100      # 100 Additive QTLs
)
```

**Default behavior:**

- Single chromosome (5 Morgan length)
- Random genotypes with allele frequency ~ Uniform(0, 1)
- 50% male, 50% female
- All individuals diploid and unrelated

:::{.callout-warning}
## Cohort Names by Sex

Remember MoBPS adds `_M` and `_F` to your cohort name by default and remember
**1 = Male** and **2 = Female** for `database` usage later. 
:::

### Controlling Allele Frequencies

Allele frequencies dramatically affect genetic architecture:

```{r}
#| eval: false

# Uniform allele frequencies (default)
pop_uniform <- creating.diploid(
  nsnp        = 1000,
  nindi       = 100,
  beta.shape1 = 1,      # Beta(1,1) = Uniform(0,1)
  beta.shape2 = 1
)

# Common variants (higher MAF)
pop_common <- creating.diploid(
  nsnp = 1000,
  nindi = 100,
  beta.shape1 = 2,      # Beta(2,2) concentrates near 0.5
  beta.shape2 = 2
)

# Rare variants (low MAF)
pop_rare <- creating.diploid(
  nsnp = 1000,
  nindi = 100,
  beta.shape1 = 0.5,    # Beta(0.5,0.5) concentrates near 0 and 1
  beta.shape2 = 0.5
)
```

![Allele frequency distributions from Beta(α, β)](https://en.wikipedia.org/wiki/Beta_distribution)

:::{.callout-tip}
## Realistic Allele Frequencies

Real populations often have many rare variants and fewer common variants. Use `beta.shape1 = 0.3, beta.shape2 = 1.5` to approximate this.
:::

### Initialization Modes

Control starting genotypes with `dataset` parameter:

```{r}
#| eval: false

# Mode 1: All zeros (000.../000...)
pop_zero <- creating.diploid(nsnp = 1000, nindi = 100,
                              dataset = "all0")

# Mode 2: Fully heterozygous (000.../111...)
pop_het <- creating.diploid(nsnp = 1000, nindi = 100,
                             dataset = "allhetero")

# Mode 3: Random (default, X₁X₂X₃.../X₄X₅X₆...)
pop_random <- creating.diploid(nsnp = 1000, nindi = 100,
                                dataset = "random")

# Mode 4: Homozygous random (X₁X₂X₃.../X₁X₂X₃... same haplotypes)
pop_homo <- creating.diploid(nsnp = 1000, nindi = 100,
                              dataset = "homorandom")
```

**When to use each mode:**

- `"random"` - General purpose, unrelated founders
- `"homorandom"` - Create inbred lines or DH lines
- `"allhetero"` - Maximum heterozygosity (F1s)
- `"all0"` - Specific starting conditions or testing

## Chromosome Structure {#sec-chromosomes}

### Single vs. Multiple Chromosomes

```{r}
#| eval: false

# Single chromosome (default)
pop_single <- creating.diploid(
  nsnp = 5000,
  chromosome.length = 5  # 5 Morgan
)

# Multiple chromosomes
pop_multi <- creating.diploid(
  nsnp = 10000,
  chr.nr = c(2000, 2000, 3000, 3000),  # SNPs per chromosome
  chromosome.length = c(2, 2, 1.5, 1.5) # Length in Morgan
)
```

**Consequences of chromosome structure:**
- More chromosomes = more recombination = faster LD decay
- Longer chromosomes = stronger linkage between distant markers
- Realistic structure improves simulation accuracy

### Using Template Species Maps

MoBPS includes common species maps:

```{r}
#| eval: false

# Cattle map
pop_cattle <- creating.diploid(
  nsnp = 50000,
  nindi = 100,
  template.chip = "cattle"   # 29 autosomes
)

# Other available templates
template.chip = "pig"       # Sus scrofa
template.chip = "chicken"   # Gallus gallus
template.chip = "sheep"     # Ovis aries
template.chip = "maize"     # Zea mays
```

These provide realistic:
- Number of chromosomes
- Relative chromosome lengths (in Morgan)
- Approximate recombination rates

:::{.callout-note}
## What Templates DON'T Include

Templates provide chromosome **structure only**, not:
- Real marker positions (SNPs are evenly spaced)
- Real allele frequencies
- Real LD patterns

For these, import real data (see next section).
:::

### Custom Genetic Maps

Provide a full genetic map:

```{r}
#| eval: false

# Create a custom map
# Columns: chr, snp_name, position_Morgan, position_bp, allele_freq
my_map <- data.frame(
  chr = c(1, 1, 1, 2, 2, 2),
  snp = c("rs001", "rs002", "rs003", "rs004", "rs005", "rs006"),
  pos_M = c(0.0, 0.5, 1.0, 0.0, 0.3, 0.6),
  pos_bp = c(1000, 50000000, 100000000, 1000, 30000000, 60000000),
  freq = c(0.2, 0.5, 0.8, 0.1, 0.45, 0.7)
)

# Use the map
population <- creating.diploid(
  map = my_map,
  nindi = 100,
  n.additive = 50
)
```

## Importing Real Data {#sec-import-real-data}

### From VCF Files

VCF (Variant Call Format) is standard for genomic data:

```{r}
#| eval: false

# Import VCF file
population <- creating.diploid(
  vcf = "path/to/genotypes.vcf",  # Or .vcf.gz
  n.additive = 100,
  vcf.maxsnp = 10000,             # Optional: limit SNPs
  vcf.maxindi = 500               # Optional: limit individuals
)
```

**What gets imported:**
- Phased or unphased genotypes
- Chromosome numbers
- Base pair positions
- Marker names (rsIDs)

### From PLINK Files

PLINK (.ped/.map or .bed/.bim/.fam) is also widely used:

```{r}
#| eval: false

# MoBPS recommends converting PLINK files to VCF format for easy import.
# Command line example (using PLINK software):
# plink --file mydata --recode vcf --out mydata

# Then import the VCF:
population <- creating.diploid(
  vcf = "mydata.vcf",
  n.additive = 100
)
```

### Converting Genotypes to Haplotypes

Imported genotype data needs proper format:

```{r}
#| eval: false

# If you have a genotype matrix (individuals × SNPs coded 0/1/2)
# Convert to haplotype format for MoBPS

# Example: genotype matrix
geno_matrix <- matrix(sample(0:2, 1000, replace = TRUE),
                      nrow = 10, ncol = 100)  # 10 indi, 100 SNPs

# Convert: each individual becomes 2 haplotypes
haplo_matrix <- matrix(0, nrow = ncol(geno_matrix), ncol = 2 * nrow(geno_matrix))

for (i in 1:nrow(geno_matrix)) {
  for (j in 1:ncol(geno_matrix)) {
    if (geno_matrix[i,j] == 0) {
      haplo_matrix[j, 2*i-1] <- 0
      haplo_matrix[j, 2*i] <- 0
    } else if (geno_matrix[i,j] == 1) {
      haplo_matrix[j, 2*i-1] <- 0
      haplo_matrix[j, 2*i] <- 1
    } else {  # == 2
      haplo_matrix[j, 2*i-1] <- 1
      haplo_matrix[j, 2*i] <- 1
    }
  }
}

# Use in MoBPS
population <- creating.diploid(
  dataset = haplo_matrix,
  n.additive = 20
)
```

## Sex Ratio Control {#sec-sex-control}

### Controlling Sex Proportions

```{r}
#| eval: false

# Equal sex ratio (default)
pop_equal <- creating.diploid(nsnp = 1000, nindi = 100,
                               sex.quota = 0.5)  # 50% female

# More females
pop_female <- creating.diploid(nsnp = 1000, nindi = 100,
                                sex.quota = 0.7)  # 70% female

# Specify exactly
pop_exact <- creating.diploid(
  nsnp = 1000,
  nindi = 100,
  sex.s = c(rep(1, 30), rep(2, 70))  # 30 males, 70 females
)
```

### One-Sex Mode

For plants or situations where sex doesn't matter:

```{r}
#| eval: false

# All individuals in same group
population <- creating.diploid(
  nsnp = 1000,
  nindi = 200,
  one.sex.mode = TRUE  # Deactivate two-sex system
)
```

## Creating Multiple Breeds/Populations {#sec-multiple-breeds}

### Sequential Addition

Create distinct founder populations:

```{r}
#| eval: false

# Create breed 1
population <- creating.diploid(
  nsnp = 5000,
  nindi = 100,
  n.additive = 100,
  founder.pool = 1,           # Mark as pool 1
  name.cohort = "Breed_A"
)

# Add breed 2 (different allele frequencies)
population <- creating.diploid(
  population = population,    # Add to existing population
  nsnp = 5000,
  nindi = 100,
  n.additive = 100,
  founder.pool = 2,           # Mark as pool 2
  name.cohort = "Breed_B",
  freq = "beta"               # Different frequencies
)
```

**Uses for founder pools:**
- Model crossbreeding programs
- Track breed composition (admixture)
- Assign breed-specific QTL effects
- Study heterosis

### Adding Chromosomes

Add additional chromosomes to existing population:

```{r}
#| eval: false

# Start with chromosome 1
pop <- creating.diploid(nsnp = 1000, nindi = 100)

# Add chromosome 2
pop <- creating.diploid(
  population = pop,
  nsnp = 1500,
  add.chromosome = TRUE  # Add, don't replace
)
```

## Marker Positions {#sec-marker-positions}

### Even Spacing (Default)

```{r}
#| eval: false

# Markers evenly distributed
pop <- creating.diploid(
  nsnp = 1000,
  chromosome.length = 5  # Spread evenly over 5M
)
```

**Best for:** Fast computation, when exact positions don't matter.

### From Base Pairs

Convert physical positions to Morgan:

```{r}
#| eval: false

# Provide base pair positions
bp_positions <- seq(1, 100000000, length.out = 5000)  # 100 Mb

pop <- creating.diploid(
  nsnp = 5000,
  bp = bp_positions,
  bpcm.conversion = 1000000   # 1 Mb = 1 cM (typical for mammals)
)
```

**Common conversion rates:**
- Mammals: 1,000,000 bp/cM (= 100,000,000 bp/Morgan)
- Chicken: 300,000 bp/cM (= 30,000,000 bp/Morgan)
- Varies by species and chromosome!

### Directly in Morgan

```{r}
#| eval: false

# Provide positions in Morgan directly
positions_M <- c(0.0, 0.01, 0.05, 0.1, 0.15, ...)  # Custom positions

pop <- creating.diploid(
  nsnp = 1000,
  snp.position = positions_M,
  position.scaling = FALSE  # Don't rescale
)
```

## Advanced Options {#sec-advanced-options}

### Genotyping Arrays

Simulate partial genotyping (chip data):

```{r}
#| eval: false

# Not all SNPs genotyped
pop <- creating.diploid(
  nsnp = 50000,          # 50K total SNPs
  nindi = 1000,
  genotyped.s = rep(c(1, 0, 0, 0), 12500),  # Every 4th SNP genotyped
  share.genotyped = 0.8  # 80% of individuals genotyped
)
```

**Uses:**
- Model cost of genotyping
- Test effects of marker density
- Simulate imputation scenarios

### Size Scaling

When founders are related (real data), scale effective size:

```{r}
#| eval: false

pop <- creating.diploid(
  vcf = "real_data.vcf",
  size.scaling = 0.7  # Effective size is 70% of actual
)
```

This affects:
- Calculations of expected inbreeding
- Expected relationships
- Effective population size estimates

## Practical Examples {#sec-examples-pop}

### Example 1: Cattle Population

```{r}
#| eval: false

# Realistic cattle breeding population
cattle <- creating.diploid(
  nsnp = 50000,
  nindi = 500,
  template.chip = "cattle",
  n.additive = 100,
  n.dominant = 20,
  beta.shape1 = 0.5,      # Some rare variants
  beta.shape2 = 1.2,
  share.genotyped = 0.8,   # 80% genotyped
  sex.quota = 0.7,         # More females (dairy)
  var.target = 100,
  name.cohort = "HolsteinFounders"
)
```

### Example 2: Maize Inbred Lines

```{r}
#| eval: false

# Fully homozygous inbred lines
maize <- creating.diploid(
  nsnp = 10000,
  nindi = 20,
  template.chip = "maize",
  dataset = "homorandom",  # Homozygous
  one.sex.mode = TRUE,      # No sexes
  n.additive = 200,
  name.cohort = "InbredLines"
)
```

### Example 3: Crossbreeding Setup

```{r}
#| eval: false

# Breed A
cross_pop <- creating.diploid(
  nsnp = 10000, nindi = 100, n.additive = 100,
  founder.pool = 1, name.cohort = "BreedA",
  beta.shape1 = 2, beta.shape2 = 2  # Common variants
)

# Breed B (different frequencies)
cross_pop <- creating.diploid(
  population = cross_pop,
  nsnp = 10000, nindi = 100, n.additive = 100,
  founder.pool = 2, name.cohort = "BreedB",
  beta.shape1 = 0.8, beta.shape2 = 2,  # Rare variants
  freq = "beta"  # Independent frequencies from Breed A
)
```

## Summary

Key concepts from this chapter:

- ✅ `creating.diploid()` initializes founder populations
- ✅ Control genomic structure (chromosomes, positions, allele frequencies)
- ✅ Import real data from VCF/PLINK or simulate data
- ✅ Use species templates for realistic chromosome structure
- ✅ Create multiple breeds with founder pools
- ✅ Control sex ratios and population composition

## What's Next?

Now that you can create populations, let's design the **trait architecture** - the genetic basis of the traits you want to select on.

Continue to [Chapter 5: Trait Architecture](05-trait-architecture.qmd)!

