# Trait Architecture {#sec-trait-architecture}

## Learning Objectives

- Design simple and complex trait architectures
- Model additive, dominance, and epistatic effects
- Create correlated traits
- Add maternal/paternal effects
- Implement breed-specific QTLs

## Overview

**Trait architecture** defines the genetic basis of traits:
- Which markers are QTLs
- Effect sizes of each QTL
- Types of genetic effects (additive, dominance, epistasis)
- Correlations between traits

MoBPS provides both **automated** and **custom** approaches.

## Automated Trait Generation {#sec-auto-traits}

### Simple Additive Trait

```{r}
#| eval: false

population <- creating.diploid(
  nsnp = 10000,
  nindi = 200,
  n.additive = 100,      # 100 additive QTLs
  var.target = 1,        # Genetic variance = 1
  mean.target = 100      # Mean breeding value = 100
)
```

**What happens:**
- 100 random SNPs chosen as QTLs
- Effects drawn from Normal(0, σ²)
- Automatically scaled to target variance

### Adding Dominance

```{r}
#| eval: false

population <- creating.diploid(
  nsnp = 10000,
  nindi = 200,
  n.additive = 100,
  n.dominant = 20,       # 20 loci with dominance
  var.additive = 0.8,    # 80% of variance from additive
  var.dominant = 0.2     # 20% from dominance
)
```

**Dominance effect modeling:**
- Random loci chosen for dominance
- Effects: genotype 0, 1, 2 each get independent effect
- Can specify `dominant.only.positive = TRUE` for heterosis modeling

### Epistatic Effects

```{r}
#| eval: false

population <- creating.diploid(
  nsnp = 10000,
  nindi = 200,
  n.additive = 80,
  n.qualitative = 10,    # Qualitative epistasis (ordered)
  n.quantitative = 5     # Quantitative epistasis (unordered)
)
```

**Epistasis types:**

1. **Qualitative:** 00 < 01 < 02 < 10 < 11 < 12 < 20 < 21 < 22 (ordered, selection always beneficial)
2. **Quantitative:** Mix of positive/negative interactions

### Effect Size Distributions

```{r}
#| eval: false

# Gaussian effects (default)
pop_gaussian <- creating.diploid(
  nsnp = 5000, nindi = 200, n.additive = 100,
  effect.distribution = "gaussian"
)

# Gamma distribution (for skewed effects)
pop_gamma <- creating.diploid(
  nsnp = 5000, nindi = 200, n.additive = 100,
  effect.distribution = "gamma",
  gamma.shape1 = 0.4,    # Shape parameter
  gamma.shape2 = 1.66    # Rate parameter
)

# Equal effects
pop_equal <- creating.diploid(
  nsnp = 5000, nindi = 200,
  n.equal.additive = 50,           # 50 QTLs
  effect.size.equal.add = 0.5      # Each effect = 0.5
)
```

## Multiple Traits {#sec-multiple-traits}

### Independent Traits

```{r}
#| eval: false

population <- creating.diploid(
  nsnp = 10000,
  nindi = 200,
  n.additive = c(100, 80, 120),    # 3 traits
  var.target = c(1, 1.5, 0.8),     # Different variances
  trait.name = c("Yield", "Quality", "Disease_Resistance")
)
```

### Correlated Traits

```{r}
#| eval: false

# Define correlation matrix
cor_matrix <- matrix(c(
  1.0,  0.5, -0.3,   # Yield
  0.5,  1.0, -0.2,   # Quality
 -0.3, -0.2,  1.0    # Disease Resistance
), nrow = 3, byrow = TRUE)

population <- creating.diploid(
  nsnp = 10000,
  nindi = 200,
  n.additive = c(100, 100, 100),
  trait.cor = cor_matrix,          # Correlations
  trait.name = c("Yield", "Quality", "Disease_Resistance")
)
```

**How it works:**
- Same markers used as QTLs for correlated traits
- Effects drawn from multivariate normal distribution
- Correlations independent of LD structure

:::{.callout-warning}
## Correlation vs. Covariance

MoBPS uses **correlation** matrices, not covariance matrices. Standardize your traits first!
:::

## Custom Trait Architecture {#sec-custom-traits}

### Single-Locus Effects (real.bv.add)

Specify exactly which SNPs are QTLs and their effects:

```{r}
#| eval: false

# Create custom effects matrix
# Columns: SNP_number, chromosome, effect_0, effect_1, effect_2
custom_effects <- rbind(
  c(120,  1, -1.0,  0.0,  1.0),   # Additive effect
  c(42,   5,  0.0,  0.0,  2.0),   # Recessive
  c(17,  22,  0.1,  0.1,  0.1)    # No effect (neutral)
)

population <- creating.diploid(
  nsnp = 10000,
  nindi = 200,
  real.bv.add = custom_effects
)
```

**Effect coding:**
- Column 1: SNP number (1 to nsnp)
- Column 2: Chromosome number
- Columns 3-5: Effects for genotypes 0, 1, 2

### Two-Locus Epistasis (real.bv.mult)

```{r}
#| eval: false

# Epistatic interaction between two loci
# 9 effects for all genotype combinations
epistatic_effects <- rbind(
  c(144, 1, 145, 1,  # SNPs 144 and 145 on chr 1
    1.0, 0.0, 0.0,   # Effects when SNP1 = 0
    0.0, 0.0, 0.0,   # Effects when SNP1 = 1
    0.0, 0.0, 0.0),  # Effects when SNP1 = 2
  c(4, 3, 188, 5,    # SNP 4 on chr 3, SNP 188 on chr 5
    0.37, 0.16, 1.35,
    1.99, 1.58, 2.51,
    0.38, 2.12, 0.98)
)

population <- creating.diploid(
  nsnp = 10000,
  nindi = 200,
  real.bv.mult = epistatic_effects
)
```

### Multi-Locus Epistasis (real.bv.dice)

For effects involving 3+ loci:

```{r}
#| eval: false

# Complex epistatic network
dice_list <- list(
  # First network: 3 SNPs
  location = list(
    matrix(c(11, 1,
             12, 1,
             16, 2), ncol = 2, byrow = TRUE)
  ),
  # Effects for all 27 combinations (3^3)
  effects = list(
    runif(27, -1, 1)  # Random effects
  )
)

population <- creating.diploid(
  nsnp = 10000,
  nindi = 200,
  real.bv.dice = dice_list
)
```

## Special Trait Types {#sec-special-traits}

### Maternal/Paternal Effects

```{r}
#| eval: false

# Trait influenced by mother's genotype
population <- creating.diploid(
  nsnp = 5000,
  nindi = 100,
  n.additive = c(100, 50),
  is.maternal = c(FALSE, TRUE),  # Trait 2 is maternal
  trait.name = c("Direct", "Maternal")
)

# Combine into single trait
population <- add.combi(
  population,
  trait = 3,
  combi.weights = c(0.7, 0.3),  # 70% direct, 30% maternal
  trait.name = "Combined"
)
```

**Use cases:**
- Birth weight (maternal nutrition)
- Milk production (maternal effects on calves)
- Egg quality (maternal contribution)

### Breed-Specific QTLs

Model QTLs that only have effects in specific breeds:

```{r}
#| eval: false

# Create two breeds
pop <- creating.diploid(
  nsnp = 5000, nindi = 100,
  founder.pool = 1,
  name.cohort = "Breed_A"
)

pop <- creating.diploid(
  population = pop, nsnp = 5000, nindi = 100,
  founder.pool = 2,
  name.cohort = "Breed_B"
)

# Add trait only active in breed A
pop <- creating.trait(
  pop,
  n.additive = 100,
  trait.pool = 1  # Only affects pool 1 (Breed A)
)
```

### Non-QTL Polygenic Traits

Simulate infinitesimal model:

```{r}
#| eval: false

population <- creating.diploid(
  nsnp = 10000,
  nindi = 200,
  n.traits = 2,              # Total traits
  n.additive = 100,          # Trait 1 has QTLs
  polygenic.variance = 1     # Trait 2 is polygenic
)
```

**Trait 2 characteristics:**
- Based on parental average + inbreeding
- No specific QTLs
- Useful for: very highly polygenic traits

## Adding Traits to Existing Populations {#sec-add-traits}

Use `creating.trait()` to add traits after population creation:

```{r}
#| eval: false

# Create population
pop <- creating.diploid(nsnp = 10000, nindi = 200)

# Later, add a new trait
pop <- creating.trait(
  pop,
  n.additive = 80,
  n.dominant = 20,
  trait.name = "NewTrait"
)
```

**Why add traits later?**
- Different traits need different architectures
- Want to replace/update trait definitions
- Building complex multi-trait models step-by-step

## Combining Traits {#sec-combine-traits}

Create composite traits from existing traits:

```{r}
#| eval: false

# Create base traits
pop <- creating.diploid(
  nsnp = 5000, nindi = 200,
  n.additive = c(100, 80, 60),
  trait.name = c("Milk", "Fat", "Protein")
)

# Create selection index
pop <- add.combi(
  pop,
  trait = 4,  # New trait number
  combi.weights = c(1, 10, 5),  # Economic weights
  trait.name = "EconomicIndex"
)
```

## Trait Standardization {#sec-standardization}

Ensure traits have desired mean and variance:

```{r}
#| eval: false

population <- bv.standardization(
  population,
  trait = 1,              # Which trait
  mean.target = 100,      # Target mean
  var.target = 1,         # Target variance
  set.zero = TRUE         # Force 0 allele effect = 0
)
```

**When to use:**
- After custom trait specification
- To compare traits on same scale
- Before economic index creation

## Practical Examples {#sec-examples-traits}

### Example 1: Dairy Cattle (Multiple Traits)

```{r}
#| eval: false

# Correlated production traits
cor_matrix <- matrix(c(
  1.0,  0.6,  0.5,   # Milk
  0.6,  1.0,  0.7,   # Fat
  0.5,  0.7,  1.0    # Protein
), nrow = 3, byrow = TRUE)

dairy <- creating.diploid(
  nsnp = 50000,
  nindi = 500,
  template.chip = "cattle",
  n.additive = c(200, 150, 150),
  n.dominant = c(20, 10, 10),
  trait.cor = cor_matrix,
  var.target = c(2000, 50, 30),
  mean.target = c(8000, 350, 320),
  trait.name = c("Milk_kg", "Fat_kg", "Protein_kg")
)
```

### Example 2: Maize Hybrid Performance

```{r}
#| eval: false

# Model heterosis with dominance
maize <- creating.diploid(
  nsnp = 10000,
  nindi = 20,
  template.chip = "maize",
  dataset = "homorandom",      # Inbred lines
  n.additive = 200,
  n.dominant = 50,
  dominant.only.positive = TRUE,  # Heterosis = positive dominance
  var.additive = 0.7,
  var.dominant = 0.3,
  trait.name = "GrainYield"
)
```

### Example 3: Disease Resistance (Binary Trait)

```{r}
#| eval: false

# Create underlying continuous trait
pop <- creating.diploid(
  nsnp = 5000, nindi = 500,
  n.additive = 50,
  mean.target = 0,
  var.target = 1,
  trait.name = "Liability"
)

# Apply threshold transformation
pop <- creating.phenotypic.transform(
  pop,
  trait = 1,
  trafo.function = function(x) { return(x > 0) }  # 50% prevalence
)
```

## Summary

- ✅ Automated trait generation for common architectures
- ✅ Custom effects for precise control
- ✅ Multiple correlated traits supported
- ✅ Dominance and epistasis modeling
- ✅ Special traits: maternal, breed-specific, polygenic
- ✅ Flexible trait combination and standardization

## What's Next?

With realistic populations and traits, let's build more complex scenarios with LD structure and data import.

Continue to [Chapter 6: Advanced Population Setup](06-advanced-population-setup.qmd)!

