# Advanced Population Setup {#sec-advanced-population-setup}

## Learning Objectives

- Generate realistic LD structure with `founder.simulation()`
- Import and use real genomic/phenotypic data
- Create populations with selection signatures
- Model population history

## Creating Realistic LD Structure {#sec-ld-structure}

Real populations have **linkage disequilibrium (LD)** from historical selection and drift. The `founder.simulation()` function creates this automatically:

```{r}
#| eval: false

# Create population with realistic LD
population <- creating.diploid(nsnp = 10000, nindi = 100)

population <- founder.simulation(
  population,
  generations = 100,        # 100 generations of random mating
  breeding.size = 200,      # Population size per generation
  mutation.rate = 0.0001    # Mutation rate per site
)
```

**What happens:**
- Simulates historical random mating
- Creates realistic LD decay with distance
- Generates allele frequency spectrum
- Produces heterozygosity patterns

### Controlling LD Levels

```{r}
#| eval: false

# Low LD (large historical population)
pop_low_ld <- founder.simulation(
  creating.diploid(nsnp = 5000, nindi = 100),
  generations = 1000,
  breeding.size = 1000
)

# High LD (bottleneck)
pop_high_ld <- founder.simulation(
  creating.diploid(nsnp = 5000, nindi = 100),
  generations = 100,
  breeding.size = 20        # Small Ne = high LD
)
```

## Modeling Selection Signatures {#sec-selection-signatures}

Simulate hard sweeps (selective sweeps):

```{r}
#| eval: false

# Create population
pop <- creating.diploid(
  nsnp = 10000,
  nindi = 200,
  chr.nr = 1,
  n.additive = 100
)

# Historical selection on trait 1
pop <- founder.simulation(
  pop,
  generations = 50,
  breeding.size = 200,
  selection.criteria = "bv",     # Select on breeding values
  selection.size = c(20, 20),    # Strong selection
  selection.m.gen = "all",
  selection.f.gen = "all"
)
```

**Creates:**
- Reduced diversity around selected QTLs
- Extended haplotype homozygosity
- Realistic post-selection LD patterns

## Importing Real Phenotypic Data {#sec-import-pheno}

Use real phenotypes to estimate trait architecture:

```{r}
#| eval: false

# Example: Load data from MoBPS example population
data(ex_pop)

# Extract information
pheno <- get.pheno(ex_pop, gen = 1:5)
geno <- get.geno(ex_pop, gen = 1:5)
map <- get.map(ex_pop, use.snp.nr = TRUE)

# Estimate marker effects from data
estimated_effects <- effect.estimate.add(geno, pheno, map)

# Create new population with estimated architecture
population <- creating.diploid(
  dataset = get.haplo(ex_pop, gen = 1),
  map = map,
  real.bv.add = estimated_effects
)
```

**Workflow:**
1. Import real genotypes + phenotypes
2. Estimate marker effects (GWAS/GBLUP)
3. Use estimated effects in simulation

## Creating Age-Structured Populations {#sec-age-structure}

Model overlapping generations:

```{r}
#| eval: false

# Create founders (year 0)
pop <- creating.diploid(
  nsnp = 5000, nindi = 100,
  n.additive = 100,
  name.cohort = "Year_0"
)

# Year 1: Some old + new animals
pop <- breeding.diploid(
  pop,
  selection.m.cohorts = "Year_0",
  selection.f.cohorts = "Year_0",
  selection.size = c(20, 40),    # Keep some old animals
  breeding.size = c(30, 60),     # Plus new calves
  name.cohort = "Year_1"
)

# Select from BOTH cohorts in Year 2
pop <- breeding.diploid(
  pop,
  selection.m.cohorts = c("Year_0", "Year_1"),
  selection.f.cohorts = c("Year_0", "Year_1"),
  selection.size = c(20, 40),
  breeding.size = c(30, 60),
  name.cohort = "Year_2"
)
```

**Applications:**
- Dairy cattle (multi-parity cows)
- Progeny testing schemes
- Conservation programs

This mirrors classical overlapping-generation breeding program formulations described by Hill [@hill1974prediction].

## Multi-Stage Pedigree Simulation {#sec-pedigree-simulation}

Create populations from pedigree structure only:

```{r}
#| eval: false

# Generate pedigree-based population
pop <- pedigree.simulation(
  population = creating.diploid(nsnp = 1000, nindi = 10),
  breeding.size = 100,
  generations = 5,
  selection.size = c(5, 10)
)
```

**Use cases:**
- Quick exploration without full genomics
- Very large pedigrees
- When genotypes not needed

## Example: Import VCF + GWAS Results {#sec-example-vcf-gwas}

Complete workflow with real data:

```{r}
#| eval: false

# Step 1: Import VCF file
pop <- creating.diploid(
  vcf = "path/to/cattle_genotypes.vcf.gz",
  vcf.maxsnp = 50000,     # Limit to 50K SNPs
  vcf.maxindi = 1000      # Limit to 1000 animals
)

# Step 2: Load phenotypes (from external file)
pheno_data <- read.csv("phenotypes.csv")
# Assume columns: AnimalID, Trait1, Trait2, Trait3

# Step 3: Match phenotypes to individuals
# (You'll need to map VCF sample IDs to phenotype IDs)

# Step 4: Estimate effects using your favorite method
# Example with rrBLUP
library(rrBLUP)
geno_matrix <- get.geno(pop, gen = 1)
effects <- mixed.solve(pheno_data$Trait1, Z = t(geno_matrix))

# Step 5: Create trait architecture from estimated effects
qtl_effects <- cbind(
  1:length(effects$u),  # SNP number
  1,                    # Chromosome (placeholder)
  -effects$u,           # Effect for allele 0
  0,                    # Effect for allele 1
  effects$u             # Effect for allele 2
)

pop <- creating.trait(
  pop,
  real.bv.add = qtl_effects,
  trait.name = "EstimatedTrait"
)

# Now ready to simulate breeding program!
```

## Summary

- ✅ `founder.simulation()` creates realistic LD structure
- ✅ Model selection signatures and population history
- ✅ Import real data (genotypes, phenotypes, GWAS results)
- ✅ Create age-structured populations
- ✅ Flexible approaches for complex scenarios

Continue to [Chapter 7: Population Examples](07-population-examples.qmd)!

