# Population Examples {#sec-population-examples}

## Learning Objectives

- See complete worked examples of population creation
- Learn patterns for common scenarios
- Understand how to combine features

## Example 1: Inbred Lines (DH Production) {#sec-example-inbred}

```{r}
#| eval: false

# Heterozygous founders
pop <- creating.diploid(
  nindi = 50,
  nsnp = 5000,
  sex.quota = 0,          # All "male" (no sex distinction)
  n.additive = 100
)

# Create DH lines (instant homozygosity)
pop <- breeding.diploid(
  pop,
  breeding.size = c(200, 0),
  dh.mating = TRUE,        # Doubled haploid technology
  name.cohort = "DH_Lines"
)

# Check homozygosity
table(get.geno(pop, gen = 2))  # Should be mostly 0 and 2, few 1s
```

**Compare with selfing:**

```{r}
#| eval: false

# 5 generations of selfing
pop_selfing <- pop
for (i in 1:5) {
  pop_selfing <- breeding.diploid(
    pop_selfing,
    breeding.size = c(200, 0),
    selfing.mating = TRUE
  )
}

# Check reduction in heterozygosity
for (gen in 1:6) {
  print(table(get.geno(pop_selfing, gen = gen)))
}
# ~50% reduction per generation
```

## Example 2: MAGIC Population (Maize) {#sec-example-magic}

Multi-parent Advanced Generation Inter-Cross:

```{r}
#| eval: false

# 20 fully homozygous founder lines
pop <- creating.diploid(
  nindi = 20,
  sex.quota = 0,
  template.chip = "maize",
  dataset = "homorandom",
  n.additive = 200
)

# Round 1: All pairwise crosses (190 F1s)
pop <- breeding.diploid(
  pop,
  breeding.size = c(190, 0),
  breeding.all.combination = TRUE,  # All combinations
  selection.size = c(20, 0),
  max.offspring = 19                # Each parent in 19 crosses
)

# Rounds 2-4: Random inter-mating
for (round in 1:3) {
  pop <- breeding.diploid(
    pop,
    breeding.size = c(190, 0),
    selection.size = c(190, 0),
    same.sex.activ = TRUE,
    same.sex.sex = 0,
    max.offspring = 2
  )
}

# Final: Create RILs by selfing
pop <- breeding.diploid(
  pop,
  breeding.size = c(380, 0),
  selfing.mating = TRUE
)
```

## Example 3: Crossbreeding Setup {#sec-example-crossbreeding}

Two-breed cross with breed-specific QTLs:

```{r}
#| eval: false

# Breed A
pop <- creating.diploid(
  nsnp = 5000, nindi = 100,
  founder.pool = 1,
  name.cohort = "Breed_A",
  n.additive = 0  # Add traits later
)

# Breed B
pop <- creating.diploid(
  population = pop,
  nsnp = 5000, nindi = 100,
  founder.pool = 2,
  name.cohort = "Breed_B",
  freq = "beta"  # Different allele frequencies
)

# Trait only in Breed A
pop <- creating.trait(pop, n.additive = 100, trait.pool = 1)

# Trait only in Breed B
pop <- creating.trait(pop, n.additive = 100, trait.pool = 2)

# Get QTL effects
qtl_eff <- get.qtl.effects(pop)

# Combine into joint trait
combined_effects <- rbind(qtl_eff[[1]][[1]], qtl_eff[[1]][[2]])

pop <- creating.trait(
  pop,
  real.bv.add = combined_effects,
  replace.traits = TRUE,
  trait.name = "Combined"
)

# Cross the breeds
pop <- breeding.diploid(
  pop,
  selection.m.cohorts = "Breed_A",
  selection.f.cohorts = "Breed_B",
  breeding.size = c(100, 100),
  name.cohort = "F1_Cross"
)

# Visualize breed composition
pools <- get.pool(pop, gen = 3, plot = TRUE)
```

## Example 4: Maternal Effects {#sec-example-maternal}

Birth weight with direct and maternal components:

```{r}
#| eval: false

# Create two traits: direct and maternal
pop <- creating.diploid(
  nsnp = 5000,
  nindi = 200,
  n.additive = c(100, 100),
  var.target = c(100, 20),
  trait.cor = matrix(c(1, 0.5, 0.5, 1), nrow = 2),
  is.maternal = c(FALSE, TRUE),
  trait.name = c("Direct", "Maternal")
)

# Combine into observed trait
pop <- add.combi(
  pop,
  trait = 3,
  combi.weights = c(1, 1),
  trait.name = "BirthWeight"
)

# Generate offspring
pop <- breeding.diploid(pop, breeding.size = c(100, 100))

# Check: maternal trait depends on dam's genotype
bvs <- get.bv(pop, gen = 2)
pedi <- get.pedigree(pop, gen = 2)
head(cbind(pedi[, 3], bvs[2, ]))  # Dam ID vs maternal BV
```

## Example 5: Import VCF + Create Trait {#sec-example-vcf-import}

```{r}
#| eval: false

# Import real genotypes
pop <- creating.diploid(
  vcf = "holstein_1000genomes.vcf.gz",
  vcf.maxsnp = 50000,
  vcf.maxindi = 500
)

# Add simulated trait
pop <- creating.trait(
  pop,
  n.additive = 200,
  n.dominant = 30,
  var.target = 100,
  mean.target = 8000,
  trait.name = "MilkYield"
)

# Generate LD structure
pop <- founder.simulation(
  pop,
  generations = 100,
  breeding.size = 300
)
```

## Example 6: Hard Sweep (Selection Signature) {#sec-example-hard-sweep}

Simulate past selection:

```{r}
#| eval: false

# Create base population
pop <- creating.diploid(
  nsnp = 10000,
  nindi = 200,
  chr.nr = 5,
  n.additive = 100,
  var.target = 1
)

# Simulate 50 generations of STRONG selection
pop <- founder.simulation(
  pop,
  generations = 50,
  breeding.size = 200,
  selection.criteria = "bv",
  selection.size = c(5, 5),      # Top 2.5%!
  phenotyping = "all",
  heritability = 0.8
)

# Now selective sweep visible in allele frequencies
qtls <- get.qtl(pop)
freq_gen1 <- get.allele.freq(pop, gen = 1, snp = qtls[[1]])
freq_gen51 <- get.allele.freq(pop, gen = 51, snp = qtls[[1]])

cbind(freq_gen1, freq_gen51)  # QTL frequencies changed
```

## Example 7: GxE (Genotype-by-Environment) {#sec-example-gxe}

```{r}
#| eval: false

# Manual approach: Create correlated "traits" for each environment
pop <- creating.diploid(
  nsnp = 5000,
  nindi = 200,
  n.additive = c(100, 100),
  trait.cor = matrix(c(1, 0.6, 0.6, 1), nrow = 2),  # Moderate correlation
  trait.name = c("Env1", "Env2")
)

# Automated GxE model
pop <- creating.diploid(
  nsnp = 5000,
  nindi = 200,
  n.additive = 100,
  gxe.model = TRUE,
  gxe.covariance = matrix(c(1, 0.6, 0.6, 1.2), nrow = 2)
)
```

## Example 8: Import PLINK Files {#sec-example-plink}

```{r}
#| eval: false

# MoBPS recommends converting PLINK files to VCF format for easy import.
# Command line example (using PLINK software):
# plink --file mydata --recode vcf --out mydata

# Then import the VCF:
pop <- creating.diploid(
  vcf = "mydata.vcf",
  n.additive = 150,
  trait.name = "ImportedTrait"
)
```

## Summary

These examples demonstrate:

- ✅ Inbred line creation (DH vs selfing)
- ✅ MAGIC population schemes
- ✅ Crossbreeding with breed-specific QTLs
- ✅ Maternal effects modeling
- ✅ VCF/PLINK data import
- ✅ Selection signatures
- ✅ Genotype-by-environment interactions

## What's Next?

With populations created, let's simulate **breeding actions** - selection, mating, and culling.

Continue to [Chapter 8: Selection Strategies](08-selection-strategies.qmd)!

