# Selection Strategies {#sec-selection-strategies}

## Overview

Selection is controlled through `breeding.diploid()` parameters. Key options:

- `selection.criteria` - What to select on
- `selection.size` - How many to select
- `selection.m.gen` / `selection.f.gen` - Which generations
- `selection.index` - Multi-trait selection

## Random Selection

```{r}
#| eval: false

pop <- breeding.diploid(
  pop,
  selection.criteria = "random",
  selection.size = c(20, 20),
  breeding.size = c(100, 100)
)
```

## Phenotypic Selection

```{r}
#| eval: false

# Phenotype first
pop <- breeding.diploid(pop, phenotyping = "all", heritability = 0.5)

# Select on phenotypes
pop <- breeding.diploid(
  pop,
  selection.criteria = "pheno",
  selection.size = c(10, 10),
  breeding.size = c(50, 50)
)
```

## Genomic Selection (BVE)

```{r}
#| eval: false

# Calculate BVEs
pop <- breeding.diploid(pop, bve = TRUE, bve.gen = 1:3)

# Select on estimated breeding values
pop <- breeding.diploid(
  pop,
  selection.criteria = "bve",
  selection.size = c(10, 10),
  breeding.size = c(50, 50)
)
```

## Selection on True Breeding Values

```{r}
#| eval: false

# Select on known BVs (simulation only!)
pop <- breeding.diploid(
  pop,
  selection.criteria = "bv",
  selection.size = c(5, 5),
  breeding.size = c(100, 100)
)
```

## Multi-Trait Selection Index

```{r}
#| eval: false

# Define economic weights
pop <- breeding.diploid(
  pop,
  selection.criteria = "bve",
  selection.index = c(1, 10, 5),  # Weights for traits 1,2,3
  selection.size = c(10, 10),
  breeding.size = c(50, 50)
)
```

## Truncation Selection with Variable Intensity

```{r}
#| eval: false

# Select top males, more females
pop <- breeding.diploid(
  pop,
  selection.criteria = "bve",
  selection.size = c(5, 20),     # 5% males, 20% females
  selection.m.gen = 4,
  selection.f.gen = 4,
  breeding.size = c(50, 50)
)
```

## Threshold Selection

```{r}
#| eval: false

# Only individuals above threshold
pop <- breeding.diploid(
  pop,
  selection.criteria = "bve",
  selection.m.threshold = 105,   # BVE > 105
  selection.f.threshold = 100,   # BVE > 100
  breeding.size = c(50, 50)
)
```

## Optimal Genetic Contribution (OGC)

Balance genetic gain with diversity:

```{r}
#| eval: false

# Requires optiSel package
library(optiSel)

pop <- breeding.diploid(
  pop,
  selection.algorithm = "ogc",
  ogc.target.inbreeding = 0.01,  # Max 1% inbreeding increase
  selection.size = c(20, 20),
  breeding.size = c(100, 100)
)
```

## Custom Selection Function

```{r}
#| eval: false

# Define custom selection
my_selector <- function(population, gen, sex) {
  bv <- get.bv(population, gen = gen, sex = sex)
  pheno <- get.pheno(population, gen = gen, sex = sex)

  # Custom criterion: BV + 0.5*pheno
  score <- bv[1,] + 0.5 * pheno[1,]

  # Return indices of selected individuals
  return(order(score, decreasing = TRUE)[1:10])
}

# Use in breeding
pop <- breeding.diploid(
  pop,
  selection.m.function = my_selector,
  selection.f.function = my_selector,
  breeding.size = c(50, 50)
)
```

## Summary

- Random, phenotypic, genomic, and true BV selection
- Multi-trait selection indices
- Threshold and OGC methods
- Custom selection functions

Continue to [Chapter 9: Mating Designs](09-mating-designs.qmd)!

